---
title: "Eviction Map Yearly Data - EDA"
author: "Alivia Castor"
date: "2024-12-04"
output: html_document
---

# RMD Notebook Research Goal

> This notebook is used to extract, combine, clean, and process the yearly eviction data sourced from ![Eviction Lab](https://evictionlab.org/). Through this notebook we aim to conduct EDA on this data to serve as a component of our research project, where we examine: **the key factors that contribute to high housing risk across boroughs, and how public health and socioeconomic indicators correlate with these risks?**


# Dataset Combination & Preprocessing

## Data Preporcessing Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(kableExtra)

```

## Combining Original DataSets 

**This data comes from yearly data sourced from `EvictionLab`**  

### Load Data and Libraries

```{r}
# Read the CSV files
queens_bronx_manhattan <- read_csv("./data/eviction_stats_Queens_Bronx_Manhattan.csv")
queens_richmond <- read_csv("./data/eviction_stats_Brooklyn_Richmond.csv")
```

### Combine `EvictionLab` Yearly Datasets

```{r}
# Combine the datasets
nyc_evictions <- bind_rows(queens_bronx_manhattan, queens_richmond) %>%
  # Remove any potential duplicates
  distinct() %>%
  # Sort by year and county name
  arrange(year, name) %>%
  # Create a borough column without "County"
  mutate(borough = str_replace(name, " County", ""))

```

### Create Data Dictionary

```{r}
# Create data dictionary
data_dictionary <- tribble(
  ~Column, ~Description,
  "GEOID", "Census FIPS code for 2010 geography",
  "name", "Census location name",
  "parent-location", "Parent location in Census hierarchy",
  "population", "Total population",
  "poverty-rate", "% of population with income below poverty level",
  "pct-renter-occupied", "% of occupied housing units that are renter-occupied",
  "median-gross-rent", "Median gross rent",
  "median-household-income", "Median household income",
  "median-property-value", "Median property value",
  "rent-burden", "Median gross rent as % of household income (max 50%)",
  "pct-white", "% population White alone, not Hispanic/Latino",
  "pct-af-am", "% population Black/African American alone, not Hispanic/Latino",
  "pct-hispanic", "% population Hispanic/Latino origin",
  "pct-am-ind", "% population American Indian/Alaska Native alone, not Hispanic/Latino",
  "pct-asian", "% population Asian alone, not Hispanic/Latino",
  "pct-nh-pi", "% population Native Hawaiian/Pacific Islander alone, not Hispanic/Latino",
  "pct-other", "% population other race alone, not Hispanic/Latino",
  "pct-multiple", "% population two or more races, not Hispanic/Latino",
  "households-threatened", "Number of unique households receiving eviction filing",
  "eviction-filings", "All eviction cases filed, including multiple cases against same address",
  "households-threatened-rate", "% of unique households receiving eviction filing",
  "eviction-filing-rate", "Ratio of evictions filed to renter-occupied homes",
  "filings-per-day", "Number of eviction filings per day"
)
```

### Summary Statistics of NYC Yearly Eviction Dataset

```{r}
# Display basic information about the dataset
cat("NYC Eviction Dataset Summary\n")
cat("==========================\n")
cat("Time period:", min(nyc_evictions$year), "-", max(nyc_evictions$year), "\n")
cat("Boroughs included:", paste(unique(nyc_evictions$borough), collapse=", "), "\n")
cat("Total records:", nrow(nyc_evictions), "\n")
cat("Number of columns:", ncol(nyc_evictions), "\n\n")

# Display first few rows of the dataset
head(nyc_evictions) %>%
  select(year, borough, population, `eviction-filings`, `households-threatened`) %>%
  kable(caption = "Sample of NYC Eviction Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Display data dictionary
data_dictionary %>%
  kable(caption = "Data Dictionary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                position = "left")

# Save combined dataset if needed
write_csv(nyc_evictions, "./data/nyc_eviction_data_combined.csv")
```

# EDA on `EvictionLab` Data

## Preprocessing 

### Load EDA Libraries

```{r}
# Load additional required packages
library(ggplot2)
library(scales)
library(gridExtra)

```

### Mapping County Names to Borough Names

```{r}
# Create simplified borough mapping
borough_mapping <- c(
  "New York" = "Manhattan (NY)",
  "Kings" = "Brooklyn (K)",
  "Queens" = "Queens (Q)",
  "Bronx" = "The Bronx (B)",
  "Richmond" = "Staten Island (R)"
)

# Update the borough names in the dataset
nyc_evictions <- nyc_evictions %>%
  mutate(borough_simple = borough_mapping[borough])

```


## Main EDA

### Temporal Trends of Evictions by Borough

```{r}
# 1. Temporal Trends of Evictions by Borough
temporal_evic_borough <- ggplot(nyc_evictions, aes(x = year, y = `eviction-filings`, color = borough_simple)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Eviction Filings Over Time by Borough",
       x = "Year",
       y = "Number of Eviction Filings",
       color = "Borough") +
  scale_y_continuous(labels = comma) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7))

temporal_evic_borough

ggsave("figures/Yearly_Eviction/temporal_evic_borough.png", plot = temporal_evic_borough, width = 8, height = 6)

```


### Relationship between Poverty Rate and Eviction Filing Rate

```{r}
# 2. Relationship between Poverty Rate and Eviction Filing Rate
poverty_evic <- ggplot(nyc_evictions, aes(x = `poverty-rate`, y = `eviction-filing-rate`, color = borough_simple)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Poverty Rate vs Eviction Filing Rate",
       x = "Poverty Rate (%)",
       y = "Eviction Filing Rate",
       color = "Borough") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))

poverty_evic

ggsave("figures/Yearly_Eviction/poverty_evic.png", plot = poverty_evic, width = 8, height = 6)

```


### Summary statistics by borough

```{r}
# 3. Summary statistics by borough
borough_summary <- nyc_evictions %>%
  group_by(borough_simple) %>%
  summarise(
    avg_eviction_rate = mean(`eviction-filing-rate`),
    avg_poverty_rate = mean(`poverty-rate`),
    avg_rent_burden = mean(`rent-burden`),
    avg_median_rent = mean(`median-gross-rent`),
    total_filings = sum(`eviction-filings`)
  ) %>%
  arrange(desc(avg_eviction_rate))

# Display summary statistics
kable(borough_summary, 
      col.names = c("Borough", "Avg Eviction Rate", "Avg Poverty Rate", 
                    "Avg Rent Burden", "Avg Median Rent", "Total Filings"),
      caption = "Borough-level Summary Statistics",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```



### Rent burden analysis

```{r}
# 4. Rent burden analysis
rent_burden <- ggplot(nyc_evictions, aes(x = borough_simple, y = `rent-burden`, fill = borough_simple)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Rent Burden Distribution by Borough",
       x = "Borough",
       y = "Rent Burden (%)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

rent_burden

ggsave("figures/Yearly_Eviction/rent_burden.png", plot = rent_burden, width = 8, height = 6)

```


### Correlation between median rent and eviction rates

```{r}
# 5. Correlation between median rent and eviction rates
med_gross_rent_evic <- ggplot(nyc_evictions, aes(x = `median-gross-rent`, y = `eviction-filing-rate`, 
                         color = borough_simple)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Median Gross Rent vs Eviction Filing Rate",
       x = "Median Gross Rent ($)",
       y = "Eviction Filing Rate",
       color = "Borough") +
  scale_x_continuous(labels = dollar) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))

med_gross_rent_evic

ggsave("figures/Yearly_Eviction/med_gross_rent_evic.png", plot = med_gross_rent_evic, width = 8, height = 6)

```


### Demographic analysis

```{r}
# 6. Demographic analysis
demographic_summary <- nyc_evictions %>%
  group_by(borough_simple) %>%
  summarise(
    white = mean(`pct-white`),
    black = mean(`pct-af-am`),
    hispanic = mean(`pct-hispanic`),
    asian = mean(`pct-asian`)
  ) %>%
  pivot_longer(cols = -borough_simple, 
               names_to = "demographic", 
               values_to = "percentage")

demog_comp_borough <- ggplot(demographic_summary, aes(x = borough_simple, y = percentage, fill = demographic)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Demographic Composition by Borough",
       x = "Borough",
       y = "Percentage",
       fill = "Demographic Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

demog_comp_borough

ggsave("figures/Yearly_Eviction/demog_comp_borough.png", plot = demog_comp_borough, width = 8, height = 6)

```


### Time series of households threatened vs actual filings

```{r}
# 7. Time series of households threatened vs actual filings
time_series <- ggplot(nyc_evictions, aes(x = year)) +
  geom_line(aes(y = `households-threatened`, color = "Households Threatened")) +
  geom_line(aes(y = `eviction-filings`, color = "Eviction Filings")) +
  facet_wrap(~borough_simple, scales = "free_y") +
  theme_minimal() +
  labs(title = "Households Threatened vs Actual Eviction Filings",
       x = "Year",
       y = "Count",
       color = "Metric") +
  scale_y_continuous(labels = comma) +
  theme(strip.text = element_text(size = 7))

time_series

ggsave("figures/Yearly_Eviction/time_series.png", plot = time_series, width = 8, height = 6)
  
```

