---
title: "Public Health Analysis for Housing Risk Scores"
author: "Alivia Castor"
date: "2024-12-04"
output: rmdformats::robobook
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Create `figures` Folder, if it doesn't exist:** 

```{r}
# Create 'figures' directory if it doesn't exist
if (!dir.exists("figures")) {
  dir.create("figures")
}
```

# Data Processing

## Load Data

```{r}
library(ggplot2)
library(dplyr)
library(readr)

# Load the cleaned dataset
data <- read_csv("data/cleaned_housing_risk_data.csv")

# Normalize the total risk score
data$total_risk_score <- data$total_risk_score / 36

str(data)

```

### Data Verification

> Data set was correctly loaded and has $59$ rows and $21$ columns\
> - The district and name columns are strings (`chr`).\
> - The remaining $19$ columns are numeric (`dbl`), including
> `covid_vax_rate`, `covid_death_rate`, `percent_uninsured`, and
> `total_risk_score`.

### Key Observations for Analysis

> **Public Health Metrics:**\
> - `covid_vax_rate:` Proportion of vaccinated individuals in each
> district.\
> - `covid_death_rate:` COVID-19 death rates in each district.\
> - `percent_uninsured:` Proportion of uninsured individuals.\
> <br> **Total Risk Score:**\
> - Already normalized (divided by $36$), making it suitable for direct
> comparisons.\
> <br> **Column Names:**\
> - Properly formatted and ready for visualization and statistical
> analysis.

<br> 
<br>

## Descriptive Statistics

```{r}
summary_stats <- data %>%
  summarise(
    mean_vax_rate = mean(covid_vax_rate, na.rm = TRUE),
    median_vax_rate = median(covid_vax_rate, na.rm = TRUE),
    var_vax_rate = var(covid_vax_rate, na.rm = TRUE),
    mean_death_rate = mean(covid_death_rate, na.rm = TRUE),
    median_death_rate = median(covid_death_rate, na.rm = TRUE),
    var_death_rate = var(covid_death_rate, na.rm = TRUE),
    mean_uninsured = mean(percent_uninsured, na.rm = TRUE),
    median_uninsured = median(percent_uninsured, na.rm = TRUE),
    var_uninsured = var(percent_uninsured, na.rm = TRUE)
  )
summary_stats

```

### Results

> -   **Mean Vaccination Rate (`mean_vax_rate`):**\
> -   Average COVID-19 vaccination rate across districts: $0.7846$
>     ($~78.46%$).\
> -   **Median Vaccination Rate (`median_vax_rate`):**\
> -   The midpoint vaccination rate: $0.775$ ($~77.5%$).\
> -   **Variance of Vaccination Rates (`var_vax_rate`):**\
> -   The spread of vaccination rates: $0.0107$.\
> -   **Mean COVID-19 Death Rate (`mean_death_rate`):**\
> -   Average death rate: $0.0052$ ($~0.52%$).\
> -   **Median COVID-19 Death Rate (`median_death_rate`):**\
> -   The midpoint death rate: $0.0053$ ($~0.53%$).

> **These descriptive statistics indicate:**\
> - Vaccination rates are relatively high and consistent across
> districts (low variance).\
> - COVID-19 death rates are low but vary slightly more than vaccination
> rates.

<br> 
<br>

# Data Visualization

## Distribution of Public Health Metrics

```{r}
# COVID-19 Vaccination Rates Distribution
vax_dist <- ggplot(data, aes(x = covid_vax_rate)) +
  geom_histogram(binwidth = 0.05, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of COVID-19 Vaccination Rates", x = "Vaccination Rate", y = "Frequency")

# Display plot inline
vax_dist

# save figures
ggsave("figures/PubHealth_OGData/vax_rate_dist.png", plot = vax_dist, width = 8, height = 6)

# COVID-19 Death Rates Distribution
death_dist <- ggplot(data, aes(x = covid_death_rate)) +  # Changed from Pdeath_dist to death_dist
  geom_histogram(binwidth = 0.001, fill = "red", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of COVID-19 Death Rates", x = "Death Rate", y = "Frequency")

# Display plot inline
death_dist  # Now matches the object name

# save figures
ggsave("figures/PubHealth_OGData/death_rate_dist.png", plot = death_dist, width = 8, height = 6)

# Percent Uninsured Distribution
uninsured_dist <- ggplot(data, aes(x = percent_uninsured)) +
  geom_histogram(binwidth = 0.01, fill = "green", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Percent Uninsured", x = "Percent Uninsured", y = "Frequency")

# Display plot inline
uninsured_dist

# save figures
ggsave("figures/PubHealth_OGData/uninsured_dist.png", plot = uninsured_dist, width = 8, height = 6)
```

### Key Observations:

> 1.  **COVID-19 Vaccination Rates**\
>
> -   The distribution shows most districts have vaccination rates
>     clustered around 75-80%, with a few outliers above $100%$.\
> -   This may indicate good overall vaccination coverage, though
>     outliers might require investigation (e.g., data entry issues or
>     over-reporting).\
>
> 2.  **COVID-19 Death Rates**\
>
> -   The majority of districts have death rates concentrated between
>     $0.004$ and $0.006$, indicating relatively low variability.\
> -   A small number of districts exhibit higher death rates, which
>     could correlate with other factors like access to healthcare or
>     socioeconomic conditions.\
>
> 3.  **Percent Uninsured**\
>
> -   Most districts have a percentage of uninsured individuals below
>     $10%$, with a few extreme outliers above $15%$.\
> -   These outliers might represent districts with limited access to
>     affordable healthcare or high poverty rates.

<br> 
<br>

## Scatter Plots of Public Health Metrics vs. Housing Risk Scores

```{r}
# Inspect column names
colnames(data)

# Vaccination Rate vs. Risk Score
vax_vs_risk <- ggplot(data, aes(x = covid_vax_rate, y = total_risk_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "blue") +
  theme_minimal() +
  labs(title = "Vaccination Rate vs Risk Score", x = "Vaccination Rate", y = "Normalized Risk Score")

# Display plot inline
vax_vs_risk

# save figures
ggsave("figures/PubHealth_OGData/vax_vs_risk.png", plot = vax_vs_risk, width = 8, height = 6)

# Death Rate vs. Risk Score
death_vs_risk <- ggplot(data, aes(x = covid_death_rate, y = total_risk_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "red") +
  theme_minimal() +
  labs(title = "Death Rate vs Risk Score", x = "Death Rate", y = "Normalized Risk Score")

# Display plot inline
death_vs_risk

# save figures
ggsave("figures/PubHealth_OGData/death_vs_risk.png", plot = death_vs_risk, width = 8, height = 6)

# Percent Uninsured vs. Risk Score
uninsured_vs_risk <- ggplot(data, aes(x = percent_uninsured, y = total_risk_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "green") +
  theme_minimal() +
  labs(title = "Percent Uninsured vs Risk Score", x = "Percent Uninsured", y = "Normalized Risk Score")

# Display plot inline
uninsured_vs_risk

# save figures
ggsave("figures/PubHealth_OGData/uninsured_vs_risk.png", plot = uninsured_vs_risk, width = 8, height = 6)
```

### Key Observations:

> 1.  **Vaccination Rate vs. Risk Score**\
>
> -   **Trend:** Negative correlation.\
> -   **Interpretation:** Districts with higher vaccination rates tend
>     to have lower housing risk scores. This suggests that vaccination
>     rates may serve as a proxy for improved public health
>     infrastructure, which reduces risk.\
>
> 2.  **Death Rate vs. Risk Score**\
>
> -   **Trend:** Positive correlation.\
> -   **Interpretation:** Districts with higher COVID-19 death rates
>     tend to have higher housing risk scores. This could indicate that
>     areas with higher mortality are more vulnerable, possibly due to
>     poor access to healthcare or underlying socioeconomic
>     disparities.\
>
> 3.  **Percent Uninsured vs. Risk Score**\
>
> -   **Trend:** Positive correlation.\
> -   **Interpretation:** A higher percentage of uninsured individuals
>     is associated with increased housing risk scores. This reinforces
>     the idea that lack of insurance is linked to vulnerability and
>     poor living conditions.

<br>
<br>

# Hypothesis Testing

## T-Test for Vaccination Rates

```{r}
median_vax_rate <- median(data$covid_vax_rate, na.rm = TRUE)
data$vax_group <- ifelse(data$covid_vax_rate > median_vax_rate, "Above Median", "Below Median")
t_test_vax <- t.test(total_risk_score ~ vax_group, data = data)
t_test_vax

```

> The Welch two-sample t-test compares the mean housing risk scores
> between districts with above-median and below-median vaccination
> rates.

### Results:

> 1.  **T-Statistic (t):**\
>
> -   $t=−4.0846$: Indicates a significant difference in means between
>     the two groups.\
>
> 2.  **Degrees of Freedom (df):**\
>
> -   $df=46.845$: Effective degrees of freedom for the test.\
>
> 3.  **P-Value:**\
>
> -   $p=0.0001712$: Extremely small p-value, which is less than the
>     typical significance level of $α=0.05$. This means the null
>     hypothesis (no difference in means) is rejected.\
>
> 4.  **Confidence Interval:**\
>
> -   $95%CI:[−0.2146,−0.0730]$: The true difference in means is
>     estimated to fall within this interval. The negative interval
>     confirms that the group with above-median vaccination rates has
>     lower risk scores.

### Group Means:

> **Above Median Vaccination Rate:**\
> - mean=$0.1790$\
> **Below Median Vaccination Rate:**\
> - mean=$0.3228$

### Interpretation:

>   
  -   Districts with higher vaccination rates have significantly lower
>     housing risk scores than those with lower vaccination rates.
> -   This suggests that higher vaccination coverage might be associated
>     with improved overall resilience and public health infrastructure,
>     reducing housing-related vulnerabilities.

<br>
<br>

## Correlation Matrix Heatmap

```{r}
library(ggplot2)
library(dplyr)

original_colnames <- colnames(data)

# Step 1: Compute correlation matrix and reshape
cor_matrix <- data %>%
  select(where(is.numeric)) %>%
  cor(method = "pearson", use = "pairwise.complete.obs")

cor_data <- as.data.frame(as.table(cor_matrix)) %>%
  rename(Var1 = Var1, Var2 = Var2, Correlation = Freq) %>%
  filter(Var1 != Var2)  # Remove diagonal (self-correlations)

# Debugging: Check the structure and head of `cor_data`
str(cor_data)
head(cor_data)

# Step 2: Filter for correlations > 0.5 (absolute value)
cor_data_filtered <- cor_data %>% filter(abs(Correlation) > 0.5)

# Debugging: Ensure filtered data is not empty
if (nrow(cor_data_filtered) == 0) {
  stop("No correlations exceed the threshold of 0.5.")
}

# Step 3: Create a ggplot heatmap
heatmap <- ggplot(cor_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", Correlation)), size = 2.5) + # Ensure two decimals
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1), space = "Lab") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8), # Adjust font size and angle
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10)
  ) +
  labs(
    title = "Correlation Heatmap of Public Health Metrics",
    subtitle = "Pearson Method, Pairwise Complete Observations",
    x = "Variables",
    y = "Variables",
    fill = "Correlation"
  )

# Save the updated plot
ggsave("figures/PubHealth_OGData/consistent_decimal_correlation_heatmap.png", plot = heatmap, width = 10, height = 8)


# Display the heatmap inline
heatmap

```


### Key Insights from the Heatmap:  

> **Strong Positive Correlations:**  
  -  **Limited English Proficiency vs. Rent Burden ($r≈0.66$):** Higher language barriers correlate with higher rent burdens.  
  -  **Crowding vs. Limited English Proficiency ($r≈0.69$):** Crowding is linked to areas with more limited English speakers.  
  -  **Housing Violations vs. Tenant Cases ($r≈0.78$):** Poor housing quality strongly correlates with tenant disputes.  
<br>
**Strong Negative Correlations:**  
  -  **Rent Burden vs. COVID Death Rate ($r≈−0.67$):** Higher rent burden correlates with lower COVID death rates, potentially reflecting urban healthcare access.  
  -  **Crowding vs. Rent Burden ($r≈−0.73$):** Crowding is more prevalent in lower-rent areas.  
<br>
**Moderate Trends:**  
  -  **Vaccination Rate vs. Risk Score ($r≈−0.38$):** Higher vaccination rates weakly correlate with lower risk scores.  
  -  **People of Color vs. Limited English Proficiency ($r≈0.56$):** Communities of color often overlap with linguistically diverse areas.  
<br>
**Weak or Non-Significant Correlations:**  
Variables like vaccination rates and eviction filings ($r≈−0.07$) show minimal connection, though this could be due to the limited data available in our initial research.  

<br>
<br>

## Bar Chart for Borough-Level Risk Scores

```{r}
# Inspect column names
colnames(data)

```

```{r}
library(stringr)
library(ggplot2)

# Check if 'district' column exists
if (!"district" %in% colnames(data)) {
  stop("The 'district' column is not found in the dataset.")
}

# Group Data by Borough
borough_comparison <- data %>%
  mutate(borough = str_extract(district, "^[A-Z]+")) %>%
  group_by(borough) %>%
  summarise(
    avg_risk_score = mean(total_risk_score, na.rm = TRUE),
    avg_vax_rate = mean(covid_vax_rate, na.rm = TRUE),
    avg_death_rate = mean(covid_death_rate, na.rm = TRUE),
    avg_uninsured = mean(percent_uninsured, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_risk_score))

# Ensure 'figures' directory exists
if (!dir.exists("figures")) {
  dir.create("figures")
}

# Bar Chart for Average Risk Scores
borough_risk_plot <- ggplot(borough_comparison, aes(x = reorder(borough, avg_risk_score), y = avg_risk_score, fill = avg_risk_score)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Risk Scores by Borough", x = "Borough", y = "Average Risk Score") +
  coord_flip()

# Save the plot
ggsave("figures/PubHealth_OGData/borough_risk_scores.png", plot = borough_risk_plot, width = 8, height = 6)

# Display the plot inline
borough_risk_plot

# Additional Visualizations: Vaccination Rate by Borough
borough_vax_plot <- ggplot(borough_comparison, aes(x = reorder(borough, avg_vax_rate), y = avg_vax_rate, fill = avg_vax_rate)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Vaccination Rates by Borough", x = "Borough", y = "Average Vaccination Rate") +
  coord_flip()

# Save the vaccination rate plot
ggsave("figures/PubHealth_OGData/borough_vaccination_rates.png", plot = borough_vax_plot, width = 8, height = 6)

# Display plot inline
borough_vax_plot

# Additional Visualizations: Death Rate by Borough
borough_death_plot <- ggplot(borough_comparison, aes(x = reorder(borough, avg_death_rate), y = avg_death_rate, fill = avg_death_rate)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Death Rates by Borough", x = "Borough", y = "Average Death Rate") +
  coord_flip()

# Save the death rate plot
ggsave("figures/PubHealth_OGData/borough_death_rates.png", plot = borough_death_plot, width = 8, height = 6)

# Display plot inline
borough_death_plot

# Additional Visualizations: Uninsured Percent by Borough
borough_uninsured_plot <- ggplot(borough_comparison, aes(x = reorder(borough, avg_uninsured), y = avg_uninsured, fill = avg_uninsured)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Average Uninsured Percent by Borough", x = "Borough", y = "Average Uninsured Percent") +
  coord_flip()

# Save the uninsured percent plot
ggsave("figures/PubHealth_OGData/borough_uninsured_percent.png", plot = borough_uninsured_plot, width = 8, height = 6)

# Display plot inline
borough_uninsured_plot

```

### Interpretation:  

>  1.  **Average Risk Scores by Borough**  
    -  **Observation:** The Bronx (BX) has the highest average risk score, followed by Brooklyn (BK), Queens (QN), Manhattan (MN), and Staten Island (SI).  
    -  **Implication:** The Bronx may have higher vulnerability related to housing and health risks compared to other boroughs. Staten Island shows the lowest risk, indicating better conditions or fewer vulnerabilities.  
  2.  **Average Vaccination Rates by Borough**  
    -  **Observation:** Manhattan (MN) has the highest average COVID-19 vaccination rates, while the Bronx (BX) and Brooklyn (BK) have the lowest.  
    -  **Implication:** Higher vaccination rates in Manhattan may correlate with better healthcare access and infrastructure. Lower rates in the Bronx and Brooklyn suggest potential public health disparities or vaccine hesitancy.  
  3.  **Average Death Rates by Borough**   
    -  **Observation:** The Bronx (BX) and Queens (QN) show the highest average death rates, while Manhattan (MN) has the lowest.  
    -  **Implication:** Higher death rates in the Bronx and Queens could reflect disparities in healthcare access, socioeconomic factors, or population density.  
  4.  **Average Uninsured Percent by Borough**   
    -  **Observation:** Queens (QN) has the highest percentage of uninsured individuals, followed by the Bronx (BX) and Brooklyn (BK). Manhattan (MN) and Staten Island (SI) have the lowest percentages.  
    -  **Implication:** A higher uninsured rate in Queens and the Bronx may contribute to greater vulnerabilities and poorer health outcomes, as access to healthcare is limited.   

<br>

### Overall Insights:  

>   
  -  The Bronx emerges as a borough with consistently higher vulnerabilities across risk scores, vaccination rates, death rates, and uninsured rates, signaling it may need targeted public health interventions.  
  -  Manhattan shows better performance across most metrics, suggesting stronger public health infrastructure and socioeconomic conditions.  
  -  Queens and Brooklyn exhibit moderate vulnerabilities, but Queens stands out for its high uninsured rate, which could impact public health outcomes.  


<br>
<br>

## ANOVA for Uninsured Percentages

```{r}
data$uninsured_group <- cut(data$percent_uninsured, breaks = 3, labels = c("Low", "Medium", "High"))
anova_uninsured <- aov(total_risk_score ~ uninsured_group, data = data)
summary(anova_uninsured)

```

### Results:

> **Degrees of Freedom (`Df`):**\
> - `uninsured_group`: $2$ groups (Low, Medium, High uninsured
> percentages).\
> - `Residuals`: $53$, indicating the remaining degrees of freedom after
> accounting for the groups.\
> **Sum of Squares (`Sum Sq`):**\
> - Variance explained by the uninsured group: $0.4326$.\
> - Residual variance (unexplained): $0.8237$.\
> **Mean Square (`Mean Sq`):**\
> - Group variance per degree of freedom: $0.21629$.\
> - Residual variance per degree of freedom: $0.01554$.\
> **F-Statistic (`F value`):**\
> - $F=13.92$, indicating that the variance between groups is much
> larger than the variance within groups.\
> **P-Value (`Pr(>F)`):**\
> - $p=1.39e−05$, highly significant ($p<0.001$).\
> - This suggests that there is a statistically significant difference
> in risk scores between the three groups of uninsured percentages.\
> **Significance Codes:**\
> - The \*\*\* next to the p-value indicates a very high level of
> statistical significance ($p<0.001$).\
> **Observations Deleted:**\
> - $3$ rows were excluded due to missing data in the percent_uninsured
> column.

### Conclusion:

> **Key Finding:**\
> The level of uninsured individuals (Low, Medium, High) significantly
> impacts the normalized housing risk scores.\
> <br> **Implication:**\
> Policies aimed at reducing the percentage of uninsured individuals may
> lower housing risk scores, particularly in districts with higher
> uninsured populations.
